---
documentclass: jss
author:
  - name: Kohei Watanabe
    orcid: 0000-0001-6519-5265
    affiliation: Waseda University
    email: \email{watanabe.kohei@gmail.com}
    url: https://posit.co
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Distributed Asymmetric Allocation of Topics for Large Imbalanced Corpora in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain: "Distributed Asymmetric Allocation of Topics for Large Imbalanced Corpora in R"
  # For running headers, if needed
  short: "Distributed Asymmetric Allocation"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
# NOTE: run rmarkdown::render("seededlda/seededlda.Rmd") in WSL to generate PDF
options(prompt = 'R> ', continue = '+ ', fig.align = 'center')
```

# Introduction

This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

$$
a = x
$$

![
Graphical model of a simple LDA. Gray circles are variables whose values are known, 
whereas white circles are latent variables whose values are unknown. $\theta$ and $\phi$ 
are the $|D| \times K$ and $K \times |V|$ matrices, respectively; $Z$ is a vector to record 
topics words, $W$, in document $d$; $d$ is one of the documents $d \in D$; $v$ is one of the 
unique words $v \in V$ in the corpus.
]("../figures/lda.png")

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.
