---
documentclass: jss
bibliography: refs.bib
author:
  - name: Kohei Watanabe
    orcid: 0000-0001-6519-5265
    affiliation: Waseda University
    address: |
      | 1-104 Totsukamachi, Shinjuku-ku, Tokyo, 169-8050, Japan
    email: \email{watanabe.kohei@gmail.com}
    url: https://github.com/koheiw/seededlda
title:
  # If you use tex in the formatted title, also supply version without
  # For running headers, if needed
  formatted: "seededlda: An \\proglang{R} package for semisupervised topic analysis of large corpora"
  plain: "seededlda: An R package for semisupervised topic analysis of large corpora"
  short: "seededlda"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [LDA, topic models, content analysis, "\\proglang{R}"]
  plain:     [LDA, topic models, content analysis, R]
preamble: >
  \usepackage{amsmath}
header-includes:
   - \usepackage{algorithm}
   - \usepackage{caption}
   - \usepackage[noend]{algpseudocode}
output: rticles::jss_article
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
# NOTE: run rmarkdown::render("seededlda/seededlda.Rmd") to generate PDF
options(prompt = 'R> ', continue = '+ ', fig.align = 'center', cache = TRUE)
```

# Introduction

Scholars have long been analyzing topics or themes of documents to understand important issues. As large textual data became more accessible, thanks to online news, social media, and digital archives, many found topic models very useful because their unsupervised algorithms allow users to identify topics in large corpora without much human involvement. Among various topic models, latent Dirichlet allocation (LDA) [@blei2003] has been arguably the most popular topic model in applied research.

Unlike latent semantic analysis (LSA) [@deerwester1990] and probabilistic LSA [@hofmann1999], LDA involves modeling the data-generating process, in which an author chooses words to create a desired mix of topics in documents. It assigns topics to individual words through iterative sampling based on co-occurrences of words in documents. This algorithm was implemented in the several R packages that include **topicmodels** [@grün2024] and **lda** [@chang2024].

The LDA was further developed to address the sparsity of word co-occurrences in short documents [@amoualian2016; @du2012; @gruber2007; @jiang2019; @yan2013; @watanabe2023], the fragmentation or agglomeration of topics [@chien2018; @syed2018; @wallach2009], and the high computational costs of iterative sampling [@newman2009; @smyth2008; @nutakki2014].

The enhanced algorithms enable LDA to become much more useful for social scientists. However, the lack of implementations in accessible software packages has led to the widespread use of the original algorithm without experiencing the advanced capabilities of the proposed alternatives.[^1] Therefore, I have developed an R package seededlda that implements algorithms for distributed computing [@newman2009], semi-supervised learning [@lu2011], sequential sampling [@enterline2000; @watanabe2023], Dirichlet prior adjustment, and convergence detection.

[^1]: For example, the *Gensim* package for Python does not support optimization of asymmetric priors when distributed computing is enabled. The *topicmodels* package for R neither performs distributed computing nor estimates asymmetric priors.

In the following sections, I first identify common problems that applied researchers face when they employ LDA in their research. Second, I explain how the enhanced algorithm solve these problems. Third, I employ the package to classify sentences from the United Nations General Assembly speech corpus [@baturo2017] to demonstrate how the algorithms lead to more plausible conclusions in content.

# Problems

Scholars often find it difficult to use LDA in their research because (1) it takes a significant amount time to identify topics in a large corpus [@newman2009]; (2) it fragments topics into sub-topics in a corpus of short documents [@nutakki2014; @lin2023]; (3) it fails to identify highly specific topics defined using seed words [@watanabe2023].

LDA takes a considerable amount of time to identify topics in a large corpus because it employs an iterative algorithm and collapsed Gibbs sampling to assign the most likely topic for each word. Furthermore, its computational cost grows proportionally to the total number of words in the corpus, the number of topics to identify, and the number of iterations [@heinrich2008]. This often leads to a quadratic increase in the execution time because the diverse content of a large corpus requires more topics to be identified. The execution time can be shortened by Gibbs sampling on multiple processors [@newman2009], although distributed computing is not very effective when the number of topics is small.

LDA tends to fragment topics into sub-topics in a corpus of short documents because it cannot accurately infer the overall probability of topics through the sampling of topics in individual documents. Usually, the frequencies of topics in long documents correlate with their overall frequencies, but they do not in short documents (e.g., sentences and social media posts) because short documents only contain words for a few related topics [@yan2013]. The use of asymmetric Dirichlet priors helps LDA to classify short documents, but it is difficult for the users to manually set the hyper-parameters.

Semi-supervised LDA fails to identify highly specific topics defined using seed words because seed words do not offer information on the frequencies of topics. Seed words increase the chance that their co-occurring words receive desired topics by inducing bias through pseudo-counts in topic assignments [@lu2011]. The use of unequal numbers of seed words (or their matches) for topics informs the Gibbs sampler about their frequencies, but the number of seed words is usually determined by the complexity or broadness of the topics instead of their frequencies [@watanabe2020].

# Algorithms

LDA infers parameters from the distributions of documents and words through Gibbs sampling. In Figure 1, the most important variables are topics, $Z_h$, and words used to express these topics, $W_h$. The Gibbs sampler iteratively assigns the most likely topics to each of the $Z_h$ based on the $\theta$ and $\phi$ distributions, whose shapes are determined by the Dirichlet priors, $\alpha_k$ and $\beta_k$, respectively. The values of the priors are equal for all the topics, $k = \{1, 2, \cdots, K\}$, in symmetric models but unequal in asymmetric models.

\begin{figure}
\includegraphics[width=10cm]{lda.png}
\caption{Graphical model of a simple LDA. Gray circles are variables whose values are known, whereas white circles are latent variables whose values are unknown. $\theta$ and $\phi$ are the $|D| \times K$ and $K \times |V|$ matrices, respectively; $Z$ is a vector to record topics words, $W$, in document $d$; $d$ is one of the documents $d \in D$; $v$ is one of the unique words $v \in V$ in the corpus.}
\end{figure}

## Basic LDA

To infer the document-topic distribution $\theta$ and the topic-word distribution $\phi$, assigned topics are saved in $M_{dk}$ and $N_{kv}$. The former is the frequency of topic k found in document $d$, the latter is the frequency of topic $k$ assigned to unique word $v$, and $\alpha_k$ and $\beta_k$ are the Dirichlet priors, which are added to the frequency counts to smooth the distributions in $\theta$ and $\phi$.[^2] The Gibbs sampler assigns topics to the words in the corpus based on the sampling distribution, $G$, derived as a product of $\theta$ and $\phi$. These relationships are defined as follows:

[^2]: For example, if $\alpha_k=0.5$, it is assumed that topic $k$ appears at least 0.5 times in any document; if $\beta_k=0.1$, it is assumed that all the words appear at least 0.1 times in topic $k$.

$$
G = P(Z=k|W=v,d)\propto\theta_{dk}\phi_{kv}
$$

$$
\theta_{dk} = P(Z=k|d,\alpha_k) = \frac{M_{dk}+\alpha_k}{M_{d.}+\sum\alpha_k}
$$ $$
\phi_{kv} = P(W=v|k,\beta_k) = \frac{N_{kv}+\beta_k}{N_{k.}+\sum\beta_k}
$$

In Algorithm 1, $z_i$ is the topic assigned for word $h$ in document $d$ by the Gibbs sampler at iteration $i$. A simple LDA completes the inference by repeating sampling of topics until $i$ reaches a fixed number of iterations, $max\_iter$, which is usually between 1000 and 3000.

\begin{algorithm}
\caption{Pseudo-code for a simple LDA. Gibbs sampling is repeated max\_iter times.}
\input{algorithm1.tex}
\end{algorithm}

## Seeded LDA

Seeded LDA guides assignment of topics using a pre-defined list words called that target topics should include. These seed words are translated into pseudo-counts, $\omega_{kv}$, and added to $N_{kv}$ in before the iterative sampling [@lu2011]. Since these pseudo counts constant, they signficicantly increase the probability that the target topic is assigned to words cooccurring with the seed words.

The size of pseudo-count is determined by a hyper-parameter $\mu$ proportional to the frequency of seed word, $f_v$, in the corpus determined. For example, the pseudo-counts added to the target topics are equal to the actual frequencies of the word in the corpus when $\mu=0.01$.

$$
{\phi}_{kv}=\frac{N_{kv} + \omega_{kv} + \beta_k}{N_{k.} + \sum{\omega_{k.}}  +\sum{\beta_k}}
$$

$$
\omega_{kv}=100\mu\frac{f_v}{\sum f}
$$

## Sequential LDA

@watanabe_seeded_2023

$$
P\left(Z=k\middle| W=v,\ d\right)\propto\theta_{dk}\theta_{\left(d-1\right)k}^\gamma\phi_{kv}\
$$

[The original algorithm was enhanced in two steps: I first implemented distributed LDA because its parallel Gibbs sampling determines how other algorithms function. Next, I added seeded LDA, sequential LDA, the Dirichlet prior optimization, and the convergence detection algorithms to the model.]{.underline} Since all these algorithms are an extension of LDA, they should be enabled or disabled independently from each other. For example, the model is distributed LDA when only the first algorithm is enabled, but it becomes a distributed seeded LDA when the second algorithm is also enabled. Therefore, the main challenge in developing the package is significantly speeding up LDA while generalizing its original algorithm for extra capabilities.

The first algorithm was initially developed for approximately distributed LDA to increase the ability of LDA to process very large corpora [@newman2009]. The second and third were for seeded sequential LDA to weakly supervise LDA using seed words in the classification of sentences [@watanabe2023]. The fourth is a low-cost algorithm for automatically adjusting Dirichlet priors to avoid fragmentation or agglomeration of topics; the fifth is a simple algorithm for convergence detection to minimize the computational costs of LDA. While the first three algorithms are adopted from earlier studies, the last two are newly developed in this study.

Following the notations in @watanabe2023, I concisely explain the algorithms with the aid of pseudo-code (Figures 2–5). The package is implemented in C++ using the Intel Thread Building Blocks library and published on CRAN.

## Distributed Computing

LDA is computationally expensive because its algorithm assigns the most likely topic, $k$, to each word in the corpus based on the sampling distribution derived from $\theta$ and $\phi$. Therefore, an analysis of a larger corpus not only increases the frequency of sampling proportionally to the total number of words in the corpus, $N$, but also the number of parameters proportionally to the number of topics, $K$, the number of documents, $D$, and the size of the vocabulary, $V$.

Among several distributed LDA algorithms, I chose the approximately distributed LDA [@newman2009] for its simplicity and generalizability. It assigns topics to words in the same way as the simple LDA, but it splits data into smaller chunks and performs Gibbs sampling on multiple processors in parallel (Figure 3). While $N_{kv}$ and $M_{dk}$ are global variables shared by all the processors, each subprocess $e=\{1,2, \cdots, E\}$ has a local variable to record the topic assignment ${\acute{N}}_{kv}^e$. In every 10 iterations, these local counts are added to the global variable, $N_{kv}$, to synchronize the topic assignment between the processors.

\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with distributed computing. Gibbs sampling is repeated max\_iter times in separate processors over $D \times batch\_size$ documents; $D$ is the total number of documents in the corpus.
}
\input{algorithm2.tex}
\end{algorithm}
\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with parallel computing and Dirichlet prior adjustment. Adjusted $\alpha_k$ serves as Dirichlet priors in the next iteration, $i+1$.
}
\input{algorithm3.tex}
\end{algorithm}

## Dirichlet Prior Optimization

LDA is often fitted with symmetric Dirichlet priors, $\alpha$ and $\beta$, because the users lack the knowledge of topic distributions. Although symmetric priors are usually inappropriate [@chien2018], the algorithm can still identify topics when the conditional and the marginal probabilities of topics are similar, $P(Z=k|d) \approx P(Z=k)$. This is often the case in a corpus of long documents, where topics can occur multiple times, but this is rarely the case in a corpus of short documents, where topics can occur only a few times.

When the condition is unsatisfied, the Gibbs sampler requires asymmetric priors for $theta$, $\{\alpha_1, \alpha_2, \cdots, \alpha_k \}$, to accurately assign topics.[^3] Since it is very difficult for the users to specify the parameters manually, I propose a new low-cost algorithm to set asymmetric priors, $\{\hat{\alpha}_1, \hat{\alpha}_2, \cdots, \hat{\alpha}_k \}$, by automatically adjusting the initial value, $\alpha$, for each topic. This allows the algorithm to update the priors for iteration i based on the posterior probability of topics in iteration $i-1$ (Figure 4).

[^3]: Following the recommendation by Wallach et al. (2009), the Dirichlet priors for $\phi$, $\beta_k$ are left symmetric in DAA.

To limit the amount of the adjustment, the algorithm computes a small constant, $\epsilon_k = \upsilon \frac{\alpha}{M_{.k}}$ , with a hyper-parameter, $0 \leq \upsilon < 1$, based on the initial random assignment of topics to words. If $\upsilon = 0$, the priors cannot change; if $\upsilon = 0.9$, $\hat{\alpha}_k$ can drop to 10% of the initial value. Each time the Gibbs sampler updates the assigned topics, $\epsilon_k$ is added to the new topic and removed from the old topic to make asymmetric priors. Since a decrease (or increase) in a prior for a topic is offset by an increase (or decrease) in priors for other topics, the total sum of the asymmetric priors remains constant.

\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with parallel computing and Dirichlet prior adjustment. Adjusted $\alpha_k$ serves as Dirichlet priors in the next iteration, $i+1$.
}
\input{algorithm3.tex}
\end{algorithm}

## Convergence Detection

In LDA, Gibbs sampling is usually repeated between 1,000 and 3,000 times because the users cannot easily determine the necessary number of iterations before fitting models. Although distributed computing helps to complete large numbers of iterations quickly, the execution time can become much shorter by only interrupting the iterations earlier when the topic assignment is stabilized.

It is difficult to detect the convergence in Gibbs sampling in general [@gelman1992], but the use of topic divergence or perplexity is infeasible because of the high computational costs. Therefore, I devised the delta statistic as a simple convergence detection criterion for DAA (Figure 5). It measures the stability of the topic assignment by comparing current and previous topics, $\delta_i = \left| z_j \ne z_{j-1} \right|$, and continues Gibbs sampling as long as the statistic is decreasing, $\delta_i \ne \delta_{i-1}$. Since the statistic tends to fall quickly in the first few hundred iterations, it can reduce the computational cost dramatically. In distributed computing, the local variable, $\acute{\delta^e} = \left| z_j \ne z_{j-1} \right|$, can be obtained in the last sub-iteration, $j=10$, and added to the global variable, $\delta_i$, to detect convergence.

\begin{algorithm}
\caption{
Pseudo-code for DAA. It enhances LDA by adding parallel computing, Dirichlet prior adjustment, and convergence detection. Convergence is checked after each sub-iteration by comparing the value of $\delta_i$ and $\delta_{i-1}$.
}
\input{algorithm4.tex}
\end{algorithm}

## Implementation

Pre-processing of textual data can be performed using **quanteda** because **seededlda** is based on the package.

```{r, message=FALSE}
library(quanteda)
library(seededlda)
library(ggplot2)

dict <- dictionary(file = "seededlda/data/seedwords.yml", levels = 2)

corp <- readRDS("seededlda/data/corpus.RDS") |>
 corpus_subset(year >= 1991)

toks <- tokens(corp, remove_punct = TRUE, remove_number = TRUE, 
               concatenator = " ") |>
 tokens_remove(stopwords("en"), min_nchar = 2, padding = TRUE) |>
 tokens_compound(dict)

dfmt_full <- dfm(toks, remove_padding = TRUE) %>% 
 dfm_trim(min_termfreq = 10)

set.seed(1234)
id <- levels(docid(dfmt_full))
id <- sample(id, length(id) * 0.1)
dfmt <- dfm_subset(dfmt_full, docid_ %in% id) 
```

### Simple LDA

```{r}
lda <- textmodel_lda(dfmt, k = 8, max_iter = 500, verbose = TRUE)
terms(lda)
```

### Distributed LDA

```{r}
lda_dis <- textmodel_lda(dfmt, k = 8, max_iter = 500, batch_size = 0.01, 
                         verbose = TRUE)
terms(lda_dis)
```

### Seeded LDA

```{r}
dict
```

```{r}
lda_seed <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
                                batch_size = 0.01, verbose = TRUE)
terms(lda_seed)
```

### Sequential LDA

```{r}
lda_seq <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
                               gamma = 0.5, batch_size = 0.01, verbose = TRUE)
terms(lda_seq)
```

```{r}
tmp <- reshape2::melt(lda_seed$theta[seq(1, 25),1:6], c("docname", "topic"))
tmp$sentence <- as.integer(stringi::stri_extract_last_regex(tmp$docname, "\\d+"))
ggplot(tmp, aes(sentence, value, color = topic)) + 
 geom_line() +
 geom_point() +
 xlab("Sentence ID") + ylab("Probability") +
    labs(group = "Topics", color = "Topics", shape = "Topics") +
    theme_minimal() +
    #facet_wrap(vars(topic)) +
    theme(legend.position = "bottom",
          axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
          axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```

### Asymmetric priors

```{r}
lda_asy <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
                               gamma = 0.5, adjust_alpha = 0.25, batch_size = 0.01, 
                               verbose = TRUE)
print(lda_asy$alpha)
```

## Convergence detection

```{r}
lda_cnv <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE, 
                               gamma = 0.5,adjust_alpha = 0.25, batch_size = 0.01, 
                               verbose = TRUE)
terms(lda_cnv)
```

## Example

```{r}
# lda_syn <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE, verbose = TRUE,
#                                batch_size = 0.01, gamma = 0.5, adjust_alpha = 0)
lda_syn <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE,
                               batch_size = 0.01, gamma = 0.5, adjust_alpha = 0,
                               verbose = TRUE)
lda_asy <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE, 
                               batch_size = 0.01, gamma = 0.5, adjust_alpha = 0.25,
                               verbose = TRUE)
```

```{r}
dat_syn <- docvars(lda_syn$data)
dat_asy <- docvars(lda_asy$data)

dat_syn$adjust_alpha <- lda_syn$adjust_alpha
dat_asy$adjust_alpha <- lda_asy$adjust_alpha

dat_syn$topic<- topics(lda_syn, select = names(lda_syn$dictionary))
dat_asy$topic <- topics(lda_asy, select = names(lda_asy$dictionary))

dat <- rbind(dat_syn, dat_asy)
dat$dummy <- rep(1, nrow(dat))
dat$adjust_alpha <- factor(dat$adjust_alpha > 0, levels = c(FALSE, TRUE),
                           labels = c("Symmetric model", "Asymmetric model"))
dat <- subset(dat, !is.na(topic))

ggplot(dat, aes(year, dummy, group = topic, color = topic, shape = topic)) +
    #geom_line() +
    stat_summary(fun = sum, geom = "line") +
    stat_summary(fun = sum, geom = "point") +
    xlab("Year") + ylab("N. of sentences") +
    labs(group = "Topics", color = "Topics", shape = "Topics") +
    theme_minimal() +
    facet_wrap(vars(adjust_alpha)) +
    theme(legend.position = "bottom",
          axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
          axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```
