---
documentclass: jss
bibliography: refs.bib
author:
  - name: Kohei Watanabe
    orcid: 0000-0001-6519-5265
    affiliation: Waseda University
    address: |
      | 1-104 Totsukamachi, Shinjuku-ku, Tokyo, 169-8050, Japan
    email: \email{watanabe.kohei@gmail.com}
    url: https://github.com/koheiw/seededlda
title:
  # If you use tex in the formatted title, also supply version without
  # For running headers, if needed
  formatted: "seededlda: An \\proglang{R} package for semisupervised topic analysis of large corpora"
  plain: "seededlda: An R package for semisupervised topic analysis of large corpora"
  short: "seededlda"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [LDA, topic models, content analysis, "\\proglang{R}"]
  plain:     [LDA, topic models, content analysis, R]
preamble: >
  \usepackage{amsmath}
header-includes:
   - \usepackage{algorithm}
   - \usepackage{caption}
   - \usepackage[noend]{algpseudocode}
output: rticles::jss_article
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
# NOTE: run rmarkdown::render("seededlda/seededlda.Rmd") to generate PDF
options(prompt = 'R> ', continue = '+ ', fig.align = 'center', cache = TRUE)
```

# Introduction

Scholars have long been analyzing topics or themes of documents to understand important issues. As large textual data became more accessible, thanks to online news, social media, and digital archives, many found topic models very useful because their unsupervised algorithms allow users to identify topics in large corpora without much human involvement. Among various topic models, latent Dirichlet allocation (LDA) [@blei2003] has been arguably the most popular topic model in applied research.

Unlike latent semantic analysis (LSA) [@deerwester1990] and probabilistic LSA [@hofmann1999], LDA involves modeling the data-generating process, in which an author chooses words to create a desired mix of topics in documents. It assigns topics to individual words through iterative sampling based on co-occurrences of words in documents. This algorithm was implemented in the several R packages that include **topicmodels** [@grün2024] and **lda** [@chang2024].

The LDA was further developed to address the sparsity of word co-occurrences in short documents [@amoualian2016; @du2012; @gruber2007; @jiang2019; @yan2013; @watanabe2023], the fragmentation or agglomeration of topics [@chien2018; @syed2018; @wallach2009], and the high computational costs of iterative sampling [@newman2009; @smyth2008; @nutakki2014].

The enhanced algorithms enable LDA to become much more useful for social scientists. However, the lack of implementations in accessible software packages has led to the widespread use of the original algorithm without experiencing the advanced capabilities of the proposed alternatives.[^1] Therefore, I have developed an R package seededlda that implements algorithms for distributed computing [@newman2009], semi-supervised learning [@lu2011], sequential sampling [@enterline2000; @watanabe2023], Dirichlet prior adjustment, and convergence detection.

[^1]: For example, the *Gensim* package for Python does not support optimization of asymmetric priors when distributed computing is enabled. The *topicmodels* package for R neither performs distributed computing nor estimates asymmetric priors.

In the following sections, I first identify common problems that applied researchers face when they employ LDA in their research. Second, I explain how the enhanced algorithm solve these problems. Third, I employ the package to classify sentences from the United Nations General Assembly speech corpus [@baturo2017] to demonstrate how the algorithms lead to more plausible conclusions in content.

# Problems

Scholars often find it difficult to use LDA in their research because (1) it takes a significant amount time to identify topics in a large corpus [@newman2009]; (2) it fragments topics into sub-topics in a corpus of short documents [@nutakki2014; @lin2023]; (3) it fails to identify highly specific topics defined using seed words [@watanabe2023].

LDA takes a considerable amount of time to identify topics in a large corpus because it employs an iterative algorithm and collapsed Gibbs sampling to assign the most likely topic for each word. Furthermore, its computational cost grows proportionally to the total number of words in the corpus, the number of topics to identify, and the number of iterations [@heinrich2008]. This often leads to a quadratic increase in the execution time because the diverse content of a large corpus requires more topics to be identified. The execution time can be shortened by Gibbs sampling on multiple processors [@newman2009], although distributed computing is not very effective when the number of topics is small.

LDA tends to fragment topics into sub-topics in a corpus of short documents because it cannot accurately infer the overall probability of topics through the sampling of topics in individual documents. Usually, the frequencies of topics in long documents correlate with their overall frequencies, but they do not in short documents (e.g., sentences and social media posts) because short documents only contain words for a few related topics [@yan2013]. The use of asymmetric Dirichlet priors helps LDA to classify short documents, but it is difficult for the users to manually set the hyper-parameters.

Semi-supervised LDA fails to identify highly specific topics defined using seed words because seed words do not offer information on the frequencies of topics. Seed words increase the chance that their co-occurring words receive desired topics by inducing bias through pseudo-counts in topic assignments [@lu2011]. The use of unequal numbers of seed words (or their matches) for topics informs the Gibbs sampler about their frequencies, but the number of seed words is usually determined by the complexity or broadness of the topics instead of their frequencies [@watanabe2020].

# Algorithms

**seededlda** implements seeded and sequential LDA with distributed computation, Dirichlet prior optimization, and convergence detection algorithms to solves aforementioned problems. The seeded, sequential and distributed algorithms are adopted from earlier studies [@lu2011; @watanabe2023; @newman2009], but the Dirichlet prior optimization and convergence detection algorithms are developed for this package.

In the following sections, each algorithm is explained using pseudo-code focusing on their differences from the basic LDA. Since all the algorithms an extension of the model based on Gibbs sampling [@heinrich2008], they can be enabled or disabled independently from each other through hyper-parameters.

## Basic LDA

LDA infers parameters from the distributions of documents and words through Gibbs sampling. In Figure 1, the most important variables are topics, $Z_h$, and words used to express these topics, $W_h$. The Gibbs sampler iteratively assigns the most likely topics to each of the $Z_h$ based on the $\theta$ and $\phi$ distributions, whose shapes are determined by the Dirichlet priors, $\alpha_k$ and $\beta_k$, respectively. The values of the priors are equal for all the topics, $k = \{1, 2, \cdots, K\}$, in symmetric models but unequal in asymmetric models.

\begin{figure}
\begin{center}
\includegraphics[width=10cm]{lda.png}
\caption{Graphical model of a simple LDA. Gray circles are variables whose values are known, whereas white circles are latent variables whose values are unknown. $\theta$ and $\phi$ are the $|D| \times K$ and $K \times |V|$ matrices, respectively; $Z$ is a vector to record topics words, $W$, in document $d$; $d$ is one of the documents $d \in D$; $v$ is one of the unique words $v \in V$ in the corpus.}
\end{center}
\end{figure}

To infer the document-topic distribution $\theta$ and the topic-word distribution $\phi$, assigned topics are saved in $M_{dk}$ and $N_{kv}$. The former is the frequency of topic $k$ found in document $d$ (Equation 1). The latter is the frequency of topic $k$ assigned to unique word $v$ (Equation 2). In both equations, $\alpha_k$ and $\beta_k$ are the Dirichlet priors, which are added to the frequency counts to smooth the distributions in $\theta$ and $\phi$.[^2] The Gibbs sampler assigns topics to the words in the corpus based on the sampling distribution, $G$, derived as a product of $\theta$ and $\phi$ (Equation 3).

[^2]: For example, if $\alpha_k=0.5$, it is assumed that topic $k$ appears at least 0.5 times in any document; if $\beta_k=0.1$, it is assumed that all the words appear at least 0.1 times in topic $k$.

\begin{align}
\theta_{dk} &= P\left(Z=k | d,\alpha_k\right) = \frac{M_{dk}+\alpha_k}{M_{d.}+\sum\alpha_k} \\
\phi_{kv} &= P\left(W=v | k,\beta_k\right) = \frac{N_{kv}+\beta_k}{N_{k.}+\sum\beta_k} \\
G &= P\left(Z=k | W=v,d\right) \propto \theta_{dk}\phi_{kv}
\end{align}

In Algorithm 1, $z_i$ is the topic assigned for word $h$ in document $d$ by the Gibbs sampler at iteration $i$. A simple LDA completes the inference by repeating sampling of topics until $i$ reaches a fixed number of iterations, $max\_iter$, which is usually between 1000 and 3000.

\begin{algorithm}
\caption{Pseudo-code for a simple LDA. Gibbs sampling is repeated max\_iter times.}
\input{algorithm1.tex}
\end{algorithm}

## Seeded LDA

Seeded LDA guides assignment of topics using a pre-defined list words called that target topics should include. These seed words are translated into pseudo-counts, $\omega_{kv}$, and added to $N_{kv}$ in before the iterative sampling [@lu2011] (Equation 4). Since these pseudo counts constant, they signficicantly increase the probability that the target topic is assigned to words cooccurring with the seed words.

The size of pseudo-count is determined by a hyper-parameter $\mu$ proportional to the frequency of seed word, $f_v$, in the corpus determined (Equation 5). For example, the pseudo-counts added to the target topics are equal to the actual frequencies of the words in the corpus when $\mu=0.01$.

\begin{align}
\phi_{kv} &= \frac{N_{kv} + \omega_{kv} + \beta_k}{N_{k.} + \sum{\omega_{k.}} +\sum{\beta_k}} \\
\omega_{kv} &= 100\mu\frac{f_v}{\sum f}
\end{align}

## Sequential LDA

Sequential LDA classifies sentences or paragraphs by modeling the transition of topics from the previous to the current document. The sampling distribution of the sequential model is is derived as the product of the current and previsou document-word distribution, $\theta_{dk}$ and $\theta_{(d-1)k}$ [@watanabe2023] (Equation 6). The the influence of the previous document is governed by a hyper-parameter $\gamma$. The previous document has the same impact as that of the current document When $\gamma=1.0$, but its impact is halved when $\gamma=0.5$. When $\gamma=0$, the sampling distribution becomes the same as the basic model.

\begin{align}
P\left(Z = k | W = v, d\right) \propto \theta_{dk} \theta_{\left(d-1\right)k} ^ \gamma\phi_{kv}
\end{align}

## Distributed Computing

LDA is computationally expensive because its algorithm assigns the most likely topic, $k$, to each word in the corpus based on the sampling distribution derived from $\theta$ and $\phi$. Therefore, an analysis of a larger corpus not only increases the frequency of sampling proportionally to the total number of words in the corpus, $N$, but also the number of parameters proportionally to the number of topics, $K$, the number of documents, $D$, and the size of the vocabulary, $V$.

Among several distributed LDA algorithms, I chose the approximately distributed LDA [@newman2009] for its simplicity and generalizability. It assigns topics to words in the same way as the simple LDA, but it splits data into smaller chunks and performs Gibbs sampling on multiple processors in parallel (Figure 3). While $N_{kv}$ and $M_{dk}$ are global variables shared by all the processors, each subprocess $e=\{1,2, \cdots, E\}$ has a local variable to record the topic assignment ${\acute{N}}_{kv}^e$. In every 10 iterations, these local counts are added to the global variable, $N_{kv}$, to synchronize the topic assignment between the processors.

\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with distributed computing. Gibbs sampling is repeated max\_iter times in separate processors over $D \times batch\_size$ documents; $D$ is the total number of documents in the corpus.
}
\input{algorithm2.tex}
\end{algorithm}
\begin{algorithm}
\caption{
Pseudo-code for LDA with parallel computing and Dirichlet prior adjustment. Adjusted $\alpha_k$ serves as Dirichlet priors in the next iteration, $i+1$.
}
\input{algorithm3.tex}
\end{algorithm}

## Dirichlet Prior Optimization

LDA is often fitted with symmetric Dirichlet priors, $\alpha$ and $\beta$, because the users lack the knowledge of topic distributions. Although symmetric priors are usually inappropriate [@chien2018], the algorithm can still identify topics when the conditional and the marginal probabilities of topics are similar, $P(Z=k|d) \approx P(Z=k)$. This is often the case in a corpus of long documents, where topics can occur multiple times, but this is rarely the case in a corpus of short documents, where topics can occur only a few times.

When the condition is unsatisfied, the Gibbs sampler requires asymmetric priors for $theta$, $\{\alpha_1, \alpha_2, \cdots, \alpha_k \}$, to accurately assign topics.[^3] Since it is very difficult for the users to specify the parameters manually, I propose a new low-cost algorithm to set asymmetric priors, $\{\hat{\alpha}_1, \hat{\alpha}_2, \cdots, \hat{\alpha}_k \}$, by automatically adjusting the initial value, $\alpha$, for each topic. This allows the algorithm to update the priors for iteration i based on the posterior probability of topics in iteration $i-1$ (Figure 4).

[^3]: Following the recommendation by Wallach et al. (2009), the Dirichlet priors for $\phi$, $\beta_k$ are left symmetric in DAA.

To limit the amount of the adjustment, the algorithm computes a small constant, $\epsilon_k = \upsilon \frac{\alpha}{M_{.k}}$ , with a hyper-parameter, $0 \leq \upsilon < 1$, based on the initial random assignment of topics to words. If $\upsilon = 0$, the priors cannot change; if $\upsilon = 0.9$, $\hat{\alpha}_k$ can drop to 10% of the initial value. Each time the Gibbs sampler updates the assigned topics, $\epsilon_k$ is added to the new topic and removed from the old topic to make asymmetric priors. Since a decrease (or increase) in a prior for a topic is offset by an increase (or decrease) in priors for other topics, the total sum of the asymmetric priors remains constant.

\begin{algorithm}
\caption{
Pseudo-code for LDA with parallel computing and Dirichlet prior adjustment. Adjusted $\alpha_k$ serves as Dirichlet priors in the next iteration, $i+1$.
}
\input{algorithm3.tex}
\end{algorithm}

## Convergence Detection

In LDA, Gibbs sampling is usually repeated between 1,000 and 3,000 times because the users cannot easily determine the necessary number of iterations before fitting models. Although distributed computing helps to complete large numbers of iterations quickly, the execution time can become much shorter by only interrupting the iterations earlier when the topic assignment is stabilized.

It is difficult to detect the convergence in Gibbs sampling in general [@gelman1992], but the use of topic divergence or perplexity is infeasible because of the high computational costs. Therefore, I devised the delta statistic as a simple convergence detection criterion for DAA (Figure 5). It measures the stability of the topic assignment by comparing current and previous topics, $\delta_i = \left| z_j \ne z_{j-1} \right|$, and continues Gibbs sampling as long as the statistic is decreasing, $\delta_i \ne \delta_{i-1}$. Since the statistic tends to fall quickly in the first few hundred iterations, it can reduce the computational cost dramatically. In distributed computing, the local variable, $\acute{\delta^e} = \left| z_j \ne z_{j-1} \right|$, can be obtained in the last sub-iteration, $j=10$, and added to the global variable, $\delta_i$, to detect convergence.

\begin{algorithm}
\caption{
Pseudo-code LDA with parallel computing, Dirichlet prior adjustment, and convergence detection. Convergence is checked after each sub-iteration by comparing the value of $\delta_i$ and $\delta_{i-1}$.
}
\input{algorithm4.tex}
\end{algorithm}

## Implementation

The package is implemented in C++ using the Intel Thread Building Blocks library for distributed computing, but its user functions are developed based on **quanteda** for flexible preprocessing of textual data. The code below analyzed the transcripts of speeches delivered at the United Nations General Assembly between 1991 and 2017 [@baturo2017] . Since the corpus contains 444,206 sentences, it poses challenges for researchers but they can produce intuitive results within reasonable time using **seededlda**.

### Data pre-processing

Pre-processing can be accomplished using **quanteda**. `tokens()` performs tokenization and feature selection such as removal of punctuation marks and numbers. `tokens_remove()` is removes common grammatical words and symbols. `tokens_compound()` combines multi-word expressions into single tokens. `dfm()` forms a document-feature matrix that will be passed to **seededlda**.

```{r, message=FALSE}
library(quanteda)
library(seededlda)
library(ggplot2)

# load seed dictionary
dict <- dictionary(file = "seededlda/data/seedwords.yml", levels = 2)

# load corpus
corp <- readRDS("seededlda/data/corpus.RDS") |>
 corpus_subset(year >= 1991)

# tokenize corpus
toks <- tokens(corp, remove_punct = TRUE, remove_number = TRUE, 
               concatenator = " ") |>
 tokens_remove(stopwords("en"), min_nchar = 2, padding = TRUE) |>
 tokens_compound(dict)

# form a document-feature matrix (DFM)
dfmt_full <- dfm(toks, remove_padding = TRUE) |>
 dfm_trim(min_termfreq = 10)

# print the DFM
print(dfmt_full)

# randomly sample 10% of speeches
set.seed(1234)
id <- levels(docid(dfmt_full))
id <- sample(id, length(id) * 0.1)
dfmt <- dfm_subset(dfmt_full, docid_ %in% id) 
```

### Simple LDA

The number of topic is specified by the user via the `k` argument. The maximum number of iteration is set to `max_iter = 500`.

```{r}
lda <- textmodel_lda(dfmt, k = 8, max_iter = 500, verbose = TRUE)
terms(lda)
```

### Distributed LDA

When `batch_size = 0.01`, the corpus is split into small batches that contain 1% of the documents and distributed across multiple processors. The package uses all the processors unless the users limits the number through `options(seededlda_threads)`. Verbose message shows that the execution time is considerable shorter than Simple LDA.

```{r}
lda_dis <- textmodel_lda(dfmt, k = 8, max_iter = 500, batch_size = 0.01, 
  verbose = TRUE)
terms(lda_dis)
```

### Seeded LDA

Topics can be pre-defined using **quanteda**'s dictionary object in `textmodel_seededlda()`. 

```{r}
dict
```

`residual` specifies the number of unseeded topics in the model.   

```{r}
lda_seed <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
  batch_size = 0.01, verbose = TRUE)
terms(lda_seed)
```

### Sequential LDA

The `gamma` exposes the $\gamma$ parameters to the users. When $gamma = 0.5$ in either `textmodel_seededlda()` or `textmodel_lda()`, the topics of preceding sentences are considered equally important as the topics of the current sentences. Seeded topics become clearer in sequential sampling is enabled.

```{r}
lda_seq <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
  gamma = 0.5, batch_size = 0.01, verbose = TRUE)
terms(lda_seq)
```

Thanks to the sequential sampling of topics, the topics change smoothly over sentences, capturing the structures of speeches.

```{r, echo=FALSE}
tmp <- reshape2::melt(lda_seed$theta[seq(1, 25),1:6], c("docname", "topic"))
tmp$sentence <- as.integer(stringi::stri_extract_last_regex(tmp$docname, "\\d+"))
ggplot(tmp, aes(sentence, value, color = topic)) + 
  geom_line() +
  geom_point() +
  xlab("Sentence ID") + ylab("Probability") +
  labs(group = "Topics", color = "Topics", shape = "Topics") +
  theme_minimal() +
  #facet_wrap(vars(topic)) +
  theme(legend.position = "bottom",
        axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```

### Asymmetric priors

Further, the $\mu$ parameter is set via the `adjust_alpha` argument. When `adjust_alpha = 0.25`, $\alpha$ can adjusted by 25% to capture more specific topics. `alpha = 0.5` for all the topics by default, but its values vary after after the adjustment.

```{r}
lda_asy <- textmodel_seededlda(dfmt, dict, residual = 2, max_iter = 500, 
  gamma = 0.5, adjust_alpha = 0.25, batch_size = 0.01, verbose = TRUE)
print(lda_asy$alpha)
```

### Convergence detection

`auto_iter = TRUE` can be used to terminate iterative sampling before it reaches `max_iter`. It often reduces the execution time significantly.

```{r}
lda_cnv <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE, 
  gamma = 0.5, adjust_alpha = 0.25, batch_size = 0.01, verbose = TRUE)
terms(lda_cnv)
```

## Example

I fitted distributed seeded LDA with symmetric or asymmetric Dirichlet priors to illustrate how they can lead to different conclusions in content analysis. Both models are fitted with the strong sequential sampling (`gamma = 0.5`) and two unseeded topics for generic words; distributed computing and convergence detection are also enabled in both models. Moreover, the Dirichlet prior adjustment is enabled in the asymmetric model (`adjust_alpha = 0.25`) but disabled (`adjust_alpha = 0`) in the symmetric model to make it equivalent to distributed LDA.

```{r}
lda_syn <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE,
  batch_size = 0.01, gamma = 0.5, adjust_alpha = 0, verbose = TRUE)
lda_asy <- textmodel_seededlda(dfmt, dict, residual = 2, auto_iter = TRUE, 
  batch_size = 0.01, gamma = 0.5, adjust_alpha = 0.25, verbose = TRUE)
```

The most frequent topic terms of the two models were very similar but ...

```{r}
terms(lda_syn)
terms(lda_asy)
```

```{r}
dat_syn <- docvars(lda_syn$data)
dat_syn$topic<- topics(lda_syn, select = names(lda_syn$dictionary))
dat_syn$type <- "symmetric"

dat_asy <- docvars(lda_asy$data)
dat_asy$topic <- topics(lda_asy, select = names(lda_asy$dictionary))
dat_asy$type <- "asymmetric"

dat <- rbind(dat_syn, dat_asy)
dat$dummy <- rep(1, nrow(dat))
dat$type <- factor(dat$type, levels = c("symmetric", "asymmetric"))
dat <- subset(dat, !is.na(topic))
```

```{r, warning=FALSE, echo=FALSE}
ggplot(dat, aes(year, dummy, group = topic, color = topic, shape = topic)) +
  stat_summary(fun = sum, geom = "line") +
  stat_summary(fun = sum, geom = "point") +
  xlab("Year") + ylab("N. of sentences") +
  labs(group = "Topics", color = "Topics", shape = "Topics") +
  theme_minimal() +
  facet_wrap(vars(type), 
             labeller = labeller(type = c(symmetric = "Symmetric model", 
                                          asymmetric = "Asymmetric model"))) +
  theme(legend.position = "bottom",
        axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))
```
