---
documentclass: jss
bibliography: refs.bib
author:
  - name: Kohei Watanabe
    orcid: 0000-0001-6519-5265
    affiliation: Waseda University
    address: |
      | 1-104 Totsukamachi, Shinjuku-ku, Tokyo, 169-8050, Japan
    email: \email{watanabe.kohei@gmail.com}
    url: https://github.com/koheiw/seededlda
title:
  formatted: "Distributed Asymmetric Allocation of Topics for Large Imbalanced Corpora in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain: "Distributed Asymmetric Allocation of Topics for Large Imbalanced Corpora in R"
  # For running headers, if needed
  short: "Distributed Asymmetric Allocation"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
header-includes:
   - \usepackage{algorithm}
   - \usepackage{caption}
   - \usepackage[noend]{algpseudocode}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
# NOTE: run rmarkdown::render("seededlda/seededlda.Rmd") in WSL to generate PDF
options(prompt = 'R> ', continue = '+ ', fig.align = 'center')
```

# Introduction

# Algorithms

## Basic LDA

To infer the document-topic distribution $\theta$ and the topic-word distribution $\phi$, assigned topics are saved in $M_{dk}$ and $N_{kv}$. The former is the frequency of topic k found in document $d$, the latter is the frequency of topic $k$ assigned to unique word $v$, and $\alpha_k$ and $\beta_k$ are the Dirichlet priors, which are added to the frequency counts to smooth the distributions in $\theta$ and $\phi$.  The Gibbs sampler assigns topics to the words in the corpus based on the sampling distribution, $G$, derived as a product of $\theta$ and $\phi$. These relationships between are defined as follows:

$$
G = P(Z=k|W=v,d)\propto\theta_{dk}\phi_{kv}
$$
$$
\theta_{dk} = P(Z=k|d,\alpha_k) = \frac{M_{dk}+\alpha_k}{M_{d.}+\sum\alpha_k}
$$
$$
\phi_{kv} = P(W=v|k,\beta_k) = \frac{N_{kv}+\beta_k}{N_{k.}+\sum\beta_k}
$$

\includegraphics[width=5]{lda.png}
\caption{Graphical model of a simple LDA. Gray circles are variables whose values are known, whereas white circles are latent variables whose values are unknown. $\theta$ and $\phi$ are the $|D| \times K$ and $K \times |V|$ matrices, respectively; $Z$ is a vector to record topics words, $W$, in document $d$; $d$ is one of the documents $d \in D$; $v$ is one of the unique words $v \in V$ in the corpus.}

\begin{algorithm}
\caption{Pseudo-code for a simple LDA. Gibbs sampling is repeated max\_iter times.}
\input{algorithm1.tex}
\end{algorithm}

```{r}
require(quanteda)
require(seededlda)
#setwd("C:/Users/watan/Dropbox/LDA2") # to delete

dict <- dictionary(file = "C:/Users/watan/Dropbox/LDA2/seedwords_glob.yml", levels = 2)

corp <- readRDS("C:/Users/watan/Dropbox/LDA2/data/corpus.RDS") |>
 corpus_subset(year >= 1991)

toks <- tokens(corp, concatenator = " ") |>
 tokens_remove("^[\\p{N}\\p{P}]+$", valuetype = "regex", padding = TRUE) %>% 
 tokens_remove(stopwords("en"),  min_nchar = 2, padding = TRUE)

dfmt <- dfm(toks, remove_padding = TRUE) %>% 
 dfm_trim(min_termfreq = 10)

dfmt0 <- dfm_sample(dfmt, 10000) 
```


```{r}
lda1 <- textmodel_lda(dfmt0, k = 6, max_iter = 1000, verbose = TRUE)
terms(lda1)
```

## Seeded LDA

@watanabe_theory-driven_2022

$$
{\phi}_{kv}=\frac{N_{kv}+\ \beta+\omega_{kv}}{N_{.v}+\beta\left|V\right|}
$$

$$
\omega_{kv}=100\mu\frac{f_v}{\sum f}
$$


```{r, eval=FALSE}
dict
lda2 <- textmodel_seededlda(dfmt0, dict, max_iter = 1000, verbose = TRUE)
terms(lda2)
```

## Sequential LDA

@watanabe_seeded_2023

$$
P\left(Z=k\middle| W=v,\ d\right)\propto\theta_{dk}\theta_{\left(d-1\right)k}^\gamma\phi_{kv}\
$$

## Distributed LDA

\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with distributed computing. Gibbs sampling is repeated max\_iter times in separate processors over $D \times batch\_size$ documents; $D$ is the total number of documents in the corpus.
}
\input{algorithm2.tex}
\end{algorithm}

\begin{algorithm}
\caption{
Pseudo-code for the enhanced LDA with parallel computing and Dirichlet prior adjustment. Adjusted $\alpha_k$ serves as Dirichlet priors in the next iteration, $i+1$.
}
\input{algorithm3.tex}
\end{algorithm}

## Distributed Asymmetric LDA

\begin{algorithm}
\caption{
Pseudo-code for DAA. It enhances LDA by adding parallel computing, Dirichlet prior adjustment, and convergence detection. Convergence is checked after each sub-iteration by comparing the value of $\delta_i$ and $\delta_{i-1}$.
}
\input{algorithm4.tex}
\end{algorithm}

